---
title: 'Peer-graded Assignment: Course Project 1'
author: "Marlon Martinez"
date: "October 29, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading and preprocessing the data

Creating a data frame with the activity.csv

```{r load}
activityDF<-read.table("activity.csv",sep = ",",header = TRUE)
head(activityDF)
```

### What is mean total number of steps taken per day?

First, it's necessary to sum all steps for each day (missing values are ingnored)

```{r}
totalByDay<-tapply(activityDF$steps,activityDF$date,FUN = sum,na.rm=TRUE)
head(totalByDay)
```

Histogram of the total number steps per day:

```{r}
hist(totalByDay,main = "Histogram of Total of Steps per Day",xlab="Steps")
  abline(v=mean(totalByDay),col="blue",lty=2)
  abline(v=median(totalByDay),col="red",lty=2)
  legend("topright",legend=c(paste("Mean:",round(mean(totalByDay))),paste("Median:",median(totalByDay))),col = c("blue","red"),lty = 2,lwd=2)

```

The mean of total steps per day is : 
```{r echo=FALSE}
mean(totalByDay)
```
and the median is: 
```{r echo=FALSE}
median(totalByDay)
```

###What is the average daily activity pattern?

Average of steps for each interval:
```{r}
avgByInterval<-tapply(activityDF$steps,activityDF$interval,FUN = mean,na.rm=TRUE)
head(avgByInterval)
```

Plotting results:

```{r}
plot(avgByInterval,type="l",xlab = "Interval position",ylab = "Average")
abline(v=which.max(avgByInterval),col="blue",lty=2)
legend("topright",legend=paste("Interval with max average: ",names(which.max(avgByInterval))),col ="blue",lty = 2,lwd=2)
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
avgByInterval[which.max(avgByInterval)]
```

### Imputing missing values

The presence of missing days may introduce bias into some calculations or summaries of the data.
Total number of missing values in the dataset:

```{r}
sum(is.na(activityDF$steps))
```

The mean for the 5-minute interval calculated before was used to fill the missing data in the dataset.
```{r}
avgDF<-data.frame(interval=as.factor(names(avgByInterval)),avg=avgByInterval)
newActivityDF<-merge(activityDF,avgDF,by="interval")
head(newActivityDF)
```

For all the missing values of the dataset of steps they will use the mean(avg) value of each interval. 
```{r}
newActivityDF$steps[is.na(newActivityDF$steps)]<-round(newActivityDF$avg[is.na(newActivityDF$steps)])
```

Sum all steps for each day (including missing values with the average of each interval)
```{r}
newTotalByDay<-tapply(newActivityDF$steps,newActivityDF$date,FUN = sum,na.rm=TRUE)
head(newTotalByDay)
```

Histogram of the total number steps per day:

```{r}
hist(newTotalByDay,main = "Histogram of Total of Steps per Day",xlab="Steps")
  abline(v=mean(newTotalByDay),col="blue",lty=2)
  abline(v=median(newTotalByDay),col="red",lty=2)
  legend("topright",legend=c(paste("Mean:",round(mean(newTotalByDay))),paste("Median:",median(newTotalByDay))),col = c("blue","red"),lty = 2,lwd=2)

```

**What is the impact of imputing missing data on the estimates of the total daily number of steps?** Now the mean and median are higher than before.
```{r}
mean(totalByDay)
mean(newTotalByDay)
```

##Are there differences in activity patterns between weekdays and weekends?

Creating two level factor variable called weekday using the dataset with missing values fixed

```{r}
newActivityDF$date<-as.POSIXct(strptime(newActivityDF$date, "%Y-%m-%d"))
newActivityDF$weekday<-"weekday"
newActivityDF$weekday[weekdays(newActivityDF$date,abbreviate = TRUE)==c("Sat","Sun")]<-"weekend"
newActivityDF$weekday<-as.factor(newActivityDF$weekday)
head(newActivityDF)
```

Creating 2 dataset with the mean of steps by interval, one for weekdays and the other for weekend

```{r}
subWeekDF<-aggregate(steps~interval,subset(newActivityDF,newActivityDF$weekday=="weekday"),FUN = mean) 
subWeekendDF<-aggregate(steps~interval,subset(newActivityDF,newActivityDF$weekday=="weekend"),FUN = mean)

subWeekDF$weekday<-"weekday"
subWeekendDF$weekday<-"weekend"
```

Merging the datasets with the average

```{r}
weekDayDF<-rbind(subWeekDF,subWeekendDF)
head(weekDayDF)
```

Plotting the mean of intervals separated by weekday

```{r}
library(ggplot2)
gp<-ggplot(data = weekDayDF,aes(interval,steps))
gp+geom_line()+facet_grid(.~weekday)+ylim(0,250)+ggtitle("Average number of steps by interval")
```

